 

### 2021-08-05 vertical-thermal-review.Rmd
incorporating Marielle & Ty's edits

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, containg trees of different sizes along various successional stages. 
Overstory trees form canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including understory microclimates by buffering extreme temperatures, solar insolation, wind, and other meteorological conditions [@zellweger_seasonal_2019]. This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

**Here, we review how the biophysical environment and plant form and function vary across the vertical canopy gradient in forests.** 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) leaf temperature ($T_{leaf}$);
(3) the leaf traits that most strongly influence $T_{leaf}$ and metabolism;
(4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) the size-structuring of whole-tree and ecosystem function.
*We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.*

@@

**The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs. `r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest.** 
In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S. National Ecological Observatory Network [NEON; Figs. `r fig_NEON_vertical`-`r SIfig_NEON_all`; Appendix `r appendix_NEON_methods`, @schimelNEONHierarchicallyDesigned2007].
While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height [@lowmanForestCanopies1995].

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six sites in the National Ecological Observatory Network (NEON)**. Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV). Vertical profiles in micrometeorological variables (d-h) at all NEON sites are shown in Fig. S1.](NEON_height_profiles/figures/profile_all.png)

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution is structured in correspondence with the distribution of tree sizes and tree architecture within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top- vs. bottom-heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns and further influence the formation of closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.

**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Figs. `r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].** 
This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than under open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs. `r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

**Wind speeds are also higher at the top of the canopy, owing to the effect of turbulence buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speeds increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to open forests and savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs. `r fig_NEON_vertical`e, `r SIfig_NEON_all`). 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the vertical profile than the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for both water and heat loss (see section "Leaf temperature").

@@

Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Figs. `r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018].
Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a]. 
In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Especially when canopy height and leaf area index (LAI) are large, maximum Tair and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].

**Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
The $T_{air}$ gradient from the canopy to the understory varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation type [@defrenneForestMicroclimatesClimate2021]. 
Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. 
Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a 'canopy greenhouse effect' [e.g., OSBS, Figs. `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

@@

**Leaf energy balance is a tradeoff between water-use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
However, when leaf water demand (influenced by VPD and photosynthetic CO~2~ demand) exceeds the rate of supply, conservation of water ($g_s$-limitation) occurs at the cost of increasing $T_{leaf}$ [Fig. `r fig_leaf_T`e, @fauset_differences_2018].
Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, during $g_s$- limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].

@@

**Following the principles outlined above, vertical microclimate profiles, leaf traits, and $g_s$ together explain *in situ* $T_{leaf}$ observations.** 
Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$,* and therefore differs little between understory and canopy.
In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies. In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs. `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020 (in review)]. 
Greater $RH$ in the inner canopy can also increase $T_{leaf}$ relative to upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].

@@

Within-canopy, leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011] at an anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. 
As much of our understanding of trait coordination is based on the leaf economics spectrum concept (@wrightWorldwideLeafEconomics2004) which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses vertically through the canopy. 
It has been established that independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy. For example, LMA, photosynthetic capacity (per area) and $g_s$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000]. Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].

@@

The vertical structure of microenvironments likely influences forest photosynthetic capacity and $T_{leaf}$ gradients [@rey-sanchez_spatial_2016] via leaf phenology. For example, deciduousness is an adaptation to conserve water during dry periods [@meakemRoleTreeSize2018], but is also structured across the vertical profile. Within tropical deciduous species, the fraction of seasonally deciduous individuals increases toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c]. Even in evergreen forests, environmental variation alters leaf age distributions [@albertCrypticPhenologyPlants2019], which affects forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and heat stress via age-specific leaf function [@albertCrypticPhenologyPlants2019]. For instance, in tropical evergreen forests, because leaf turnover rates of sun leaves in the upper canopy is faster than in the subcanopy, less damage accumulates [@millerOnlySunlitLeaves2021]. Leaf age affects vulnerability to heat stress and has a strong effect on metabolism  [@mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015; @kikuzawaSynthesisRelationshipsLeaf2006], and can be expected to similarly scale through vertically structured phenology to affect forest responses to heat anomalies.

@@

Maximum $g_s$ increases with light environment, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table 2). However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.  $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. In open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer duration than upper canopy leaves, resulting in elevated $T_{leaf}$ [Fig. `r fig_leaf_T`,@zweifelMiddayStomatalClosure2002]. The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].

>>>>

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, containg trees of different sizes along various successional stages. 
Overstory trees form canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including understory microclimates by buffering extreme temperatures, solar insolation, wind, and other meteorological conditions [@zellweger_seasonal_2019]. This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

**Here, we review how the biophysical environment and plant form and function vary across the vertical canopy gradient in forests.** 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) leaf temperature ($T_{leaf}$);
(3) the leaf traits that most strongly influence $T_{leaf}$ and metabolism;
(4) leaf metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) the size-structuring of whole-tree and ecosystem function.
*We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.*

@@

**The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs. `r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest.** 
In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from focal sites within the U.S. National Ecological Observatory Network [NEON; Figs. `r fig_NEON_vertical`-`r SIfig_NEON_all`; Supporting Information Methods `r appendix_NEON_methods`, @schimelNEONHierarchicallyDesigned2007].
While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height [@lowmanForestCanopies1995].

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six sites six US forest sites with contrasting forest structures in the National Ecological Observatory Network (NEON)**. Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV). Vertical profiles in micrometeorological variables (d-h) at all NEON sites are shown in Fig. S1.](NEON_height_profiles/figures/profile_all.png)

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution is structured in correspondence with the distribution of tree sizes and tree architecture within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top- vs. bottom-heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns and further influence the formation of closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @smithSeasonalDroughtrelatedChanges2019; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Figs. `r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].** 
This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs. `r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

**Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].**
Wind speeds increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs. `r fig_NEON_vertical`e, `r SIfig_NEON_all`). 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the vertical profile than the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for both water and heat loss (see section "Leaf temperature").

@@

Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Figs. `r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018].
Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a]. 
In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].

**Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation type [@defrenneForestMicroclimatesClimate2021]. 
Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. 
Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a 'canopy greenhouse effect' [e.g., OSBS, Figs. `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

@@

**Leaf energy balance is a tradeoff between water-use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
However, when leaf water demand (influenced by VPD and photosynthetic CO~2~ demand) exceeds the rate of supply, conservation of water ($g_s$-limitation) occurs at the cost of increasing $T_{leaf}$ [Fig. `r fig_leaf_T`e, @fauset_differences_2018].
Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, but during $g_s$-limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].

@@

**Following the principles outlined above, vertical microclimate profiles, leaf traits, and $g_s$ together explain *in situ* $T_{leaf}$ observations.** 
Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$,* and therefore differs little between understory and canopy.
In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopy than buffered lower-canopies. In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs. `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020 (in review)]. 
Greater $RH$ in the inner canopy can also increase $T_{leaf}$ relative to upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].

@@

Within-canopy, leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011] at an anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. 
As much of our understanding of trait coordination is based on the leaf economics spectrum concept [@wrightWorldwideLeafEconomics2004] which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses vertically through the canopy. 
It has been established that independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy. For example, leaf mass per area (LMA), photosynthetic capacity ($A_A$) and $g_s$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000]. Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].

@@

The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology. For example, deciduousness is an adaptation to conserve water during dry periods [@meakemRoleTreeSize2018], but is also structured across the vertical profile. Within tropical deciduous species, the fraction of seasonally deciduous individuals increases toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c]. Even in evergreen forests, environmental variation alters leaf age distributions [@albertCrypticPhenologyPlants2019], which affects forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and heat stress via age-specific leaf function [@albertCrypticPhenologyPlants2019]. For instance, in tropical evergreen forests, because leaf turnover rates of sun leaves in the upper canopy is faster than in the subcanopy, less damage accumulates [@millerOnlySunlitLeaves2021]. Leaf age affects vulnerability to heat stress and has a strong effect on metabolism  [@mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015; @kikuzawaSynthesisRelationshipsLeaf2006], and can be expected to similarly scale through vertically structured phenology to affect forest responses to heat anomalies.

@@

Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table 2). However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.  $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. In open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer duration than upper canopy leaves, resulting in elevated $T_{leaf}$ [OSBS, Fig. `r fig_leaf_T`,@zweifelMiddayStomatalClosure2002]. The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
